Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-04 source/paper.tex
proper terminator toy

Compatible finite element methods provide a route to stable,
consistent discretisations of the partial differential equations
governing geophysical fluids \citep{cotter2023compatible}. They
possess many of the desirable properties described by
\citep{staniforth2012horizontal} as essential for the accuracy of the
dynamical core of weather and climate models, such as conservation of
mass (and potentially other quantities such as energy and enstrophy),
correct wave propagation speeds and avoidance of spurious
computational modes, even on non-orthogonal grids. Obtaining these
properties on a non-orthogonal grid is important because weather and
climate models achieve optimal performance on supercomputers by
performing calculations concurrently, or in parallel. In practice,
this means that the spatial domain is partitioned into subdomains that
are shared across the processors and data on the boundaries of the
subdomains needs to be shared between the processors to compute a
global solution. Current trends in supercomputer architectures

are run on massivly parallel supercomputers that require the co Spatial domain decomposition methods share the computation
across processors by While the required properties of the
discretisation guide the choice of finite element spaces, there are
many options for both the spatial and temporal discretisation and it
is not clear which will be best for any particular problem. Gusto
provides a flexible, extensible, easy-to-use toolkit of options for
rapid prototyping of different spatial and temporal discretisations
based on the compatible finite element framework applied to
geophysical fluid dynamics equations relevant to numerical weather and
climate prediction. In this paper we introduce Gusto to the community,
summarise its current capabilities and outline our vision for future
development.

@@

The finite element method represents fields (e.g. velocity, pressure,
density or temperature) on a mesh made of non-overlapping, conforming
cells, or elements. Within each element, the field is represented as a
sum of coefficients multiplied by polynomial basis functions. The
finite element space is specified by the degree of the polynomial
basis functions and the continuity of these functions between the
elements. The partial differential equations used in weather and
climate modelling have multiple prognostic fields and it is not
necessary, or indeed desirable, to use the same finite element
representation for each field. Using different finite element
representations for each field gives what is known as a mixed finite
element discretisation.

`Compatible' or `mimetic' finite element discretisations are
constructed by choosing the different finite element representations
so that discrete versions of continuous vector calculus identities
hold. These methods are useful in a variety of applications due to
their ability to capture the underlying geometric structure of the
governing equations. In the context of geophysical fluid dynamics,
these methods are useful because they represent geostrophic balance
exactly which is crucial for reproducing accurate simulations of the
large-scale flow. In addition to this, there is flexibility within the
compatible finite element framework to choose representations that
also give other advantages such as energy conservation, local mass
conservation and discrete dispersion properties that lead to no
spurious pressure modes.

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernel of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernel of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
onto the kernel of $\nabla\cdot$, ensuring that
$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
Rham complex such that instead $\nabla$ maps onto the kernel of
$\nabla^\perp$, ensuring that $\nabla^\perp\cdot\nabla(\cdot)=0$, but
the former is more useful in constructing discretisations for
geophysical fluid dynamics since it enables schemes with local mass
conservation \citep{}.

@@

Due to these promising properties, the next Met Office dynamical core
will use the lowest order compatible finite element discretisation on
the cubed sphere \citep{}. However, there are many important avenues
for research and unanswered questions relating to, for example,
timestepping methods, linear solver techniques and how to couple
unresolved physics processes to the resolved dynamics. This type of
research is difficult in operational models, which are generally
designed to execute specific configurations very efficiently. The
purpose of Gusto is to facilitate this research by providing an open
source, flexible, extensible framework

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similarities between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport, diffusion) seen in section
\ref{sec: governing}. In section \ref{sec: design} we describe the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.

@@

The finite element discretisation is formed by choosing appropriate
representations, or function spaces, for the prognostic variables,
multiplying by test functions from these function spaces and
integrating over the domain. In addition to this, we integrate by
parts when the inter-element continuity of the basis functions is not
sufficient to define the result of applying the differential operator
directly.

@@

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable. The spatial order of the
finite element spaces also imposes requirements on the choice of
transport scheme. In particular, when the lowest order space
$\mathbb{V}_p$ is chosen to be DG0, we need to use the recovered
transport scheme described in \citet{bendall2019recovered} to achieve
second order convergence overall. Here we describe the default
transport schemes used in Gusto for higher order function spaces and
refer the reader to \citet{bendall2019recovered} for details of the
recovered schemes, which are also implemented in Gusto.

These schemes do not preserve the monotonicity of the continuous
equation, generating spurious over and undershoots. When monotonicity
is important, for example, to avoid generating spurious and unphysical
negative values of moisture variables, we can apply a limiter as part
of the transport scheme.

\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}

@@

\section{Results}
\label{sec: results}
The aim of this section is to present a selection of results that
demonstrate the tools available in Gusto, such as the choice of grids,
HDiv element families, order of finite element spaces, different
spatial discretisation options, timestepping schemes and physics. The
numerical properties of these configurations (convergence, stability,
accuracy and efficiency) have either been demonstrated elsewhere, or
will be investigated more thoroughly in future publications.

@@

\subsection{Terminator `toy' chemistry}
This test combines transport and a `toy' chemistry model to
demonstrate physics-dynamics coupling in Gusto (where here chemistry
is the `physics').

@@

\subsubsection{Held Suarez}

@@

\appendix
\section{}    %% Appendix A

>>>>

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}. Achieving these
properties on a non-orthogonal grid is important because of parallel
scaling**. While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

@@

The finite element method represents fields (e.g. velocity, pressure,
density or temperature) on a mesh made of non-overlapping, conforming
cells, or elements. Within each element, the field is represented as a
sum of coefficients multiplied by polynomial basis functions. The
finite element space is specified by the degree of the polynomial
basis functions and the continuity of these functions between the
elements. The partial differential equations used in weather and
climate modelling have multiple prognostic fields and it is not
necessary, or indeed desirable, to use the same finite element
function space for each field. Using different finite element function
spaces for each field gives what is known as a mixed finite element
discretisation.

`Compatible' or `mimetic' finite element discretisations are
constructed by choosing the different finite element function spaces
so that discrete versions of continuous vector calculus identities
hold. These methods are useful in a variety of applications due to
their ability to capture the underlying geometric structure of the
governing equations. In the context of geophysical fluid dynamics,
these methods are useful because they represent geostrophic balance
exactly which is crucial for reproducing accurate simulations of the
large-scale flow. In addition to this, there is flexibility within the
compatible finite element framework to choose function spaces that
also give other advantages such as energy conservation, local mass
conservation and discrete dispersion properties that lead to no
spurious pressure modes.

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernal of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernal of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernal
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
onto the kernal of $\nabla\cdot$, ensuring that
$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
Rham complex such that instead we have that $\nabla$ maps onto the
kernal of $\nabla^\perp$, ensuring that
$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
constructing discretisations for geophysical fluid dynamics since it
enables schemes with local mass conservation \citep{}.

@@

Given that there are many promising avenues for research in the area
of compatible finite element discretisations for geophysical fluid
dynamics (**where to say what these are?), a flexible, extensible
implementation of the methods is strongly desirable. This is
especially required because, while it is feasible for someone to write
a code implementing a specific finite element discretisation of a PDE,
the code quickly becomes complicated and unsustainable when
investigating different orders or families of finite elements, or
different formulations of the equation. For geophysical fluid dynamics
applications there are additional considerations relating to the
geometry and types of waves present in the flow. For example, the
scales of motion in the horizontal and vertical directions differ by
several orders of magnitude. (also: boundaries, vertical staggering
for theta space)

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similaries between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport, diffusion) seen in section
\ref{sec: governing}. In section \ref{sec: design} we describe the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.

@@

The finite element discretisation is formed by choosing appropriate
function spaces for the prognostic variables, multiplying by test
functions from these function spaces and integrating over the
domain. In addition to this, we integrate by parts when the
inter-element continuity of the basis functions is not sufficient to
define the result of applying the differential operator directly.

@@

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable.

@@

\section{Results}
\label{sec: results}
The aim of this section is to present a selection of results that
showcase the tools available in Gusto, such as the choice of grids,
HDiv element families, order of finite element spaces, different
spatial discretisation options and timestepping schemes.
As these options can be combined in so many ways, it is only possible to present results from a tiny fraction configurations.
We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation. Such results are demonstrated elsewhere, for instance \cite{bendall2020compatible} or \cite{hartney2025exploring}.
The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.

@@

\subsection{Terminator Toy chemistry}
The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
The chemicals interact such that the sum $2m_X+m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms.

@@

@@

\appendix
\section{Test Case Settings}  \label{sec:appendix}  %% Appendix A
